Bayesian Non-parametric Inference for Multivariate\\ Peaks-over-threshold Models

### 2024-02-01 conclusion.tex
changes

\section{Conclusion\label{sec:conclusion}}
In this paper, we have built upon the definition of the multivariate Pareto 
    described in \cite{ferreira2014} to establish a useful representation of 
    its dependence structure, through the distribution of its angular component,
    which is supported on the positive orthant of the unit hypersphere under 
    the $\mathcal{L}_{\infty}$ norm.  Due to the inherent difficulty of 
    obtaining the likelihood of distributions with support on ${\mathbb S}^{d-1}_\infty$ 
    our method transforms the data to ${\mathbb S}_{p}^{d-1}$, fits then using mixtures of 
    products of independent gammas, then transforms the predictions back to 
    ${\mathbb S}^{d-1}_\infty$. As ${\mathbb S}_{p}^{d-1}$ converges to 
    ${\mathbb S}_{\infty}^{d-1}$ as  $p\to\infty$, we expect the proposed 
    resampling to be efficient for large enough $p$. In fact, our exploration 
    of the simulated and real data indicates that the procedure is robust to 
    the choice of moderately large values of $p$. 
    Our method includes two inferential steps. The first consists of the
    estimation of the marginal Pareto distributions. The second consists 
    of the estimation of the angular density. Parameter uncertainty incurred in
    the former is not propagated to the latter. Conceptually, an integrated approach 
    that accounts for all the estimation uncertainty is conceivable. Unfortunately, 
    this leads to posterior distributions with complex data dependent restrictions 
    that  are very difficult to explore, especially in large dimensional settings. In 
    fact, our attempts to fit a simple parametric model for the marginals and the
    angular measure jointly in several dimensions were not successful.

@@

>>>>

\section{Conclusion\label{sec:conclusion}}
In this paper, we have built upon the definition of the multivariate Pareto 
    described in \cite{ferreira2014} to establish a useful representation of 
    its dependence structure through the distribution of its angular component,
    which is supported on the positive orthant of the unit hypersphere under 
    the $\mathcal{L}_{\infty}$ norm, $\mathbb{S}_{\infty}^{d-1}$.
    Due to the inherent difficulty of 
    obtaining the likelihood of distributions with support on ${\mathbb S}^{d-1}_\infty$ 
    our method transforms the data to ${\mathbb S}_{p}^{d-1}$, fits then using mixtures of 
    products of independent gammas, then transforms the predictions back to 
    ${\mathbb S}^{d-1}_\infty$. As ${\mathbb S}_{p}^{d-1}$ converges to 
    ${\mathbb S}_{\infty}^{d-1}$ as  $p\to\infty$, we expect the proposed 
    resampling to be efficient for large enough $p$. In fact, our exploration 
    of the simulated and real data indicates that the procedure is robust to 
    the choice of moderately large values of $p$.
    Our method includes two inferential steps. The first consists of the
    estimation of the marginal Pareto distributions; the second consists 
    of the estimation of the angular density. Parameter uncertainty incurred in
    the former is not propagated to the latter. Conceptually, an integrated approach 
    that accounts for all the estimation uncertainty is conceivable. Unfortunately, 
    this leads to posterior distributions with complex data dependent restrictions 
    that are very difficult to explore, especially in large dimensional settings. In 
    fact, our attempts to fit a simple parametric model for the marginals and the
    angular measure jointly in several dimensions were not successful.

@@

Establishing the angular measure
    specifically in $\mathbb{S}_{\infty}^{d-1}$ provides two benefits over 
    $\mathbb{S}_p^{d-1}$.  First, the transformation to 
    $\mathbb{S}_{\infty}^{d-1}$ is unique.  Observe in 
    Equation~\eqref{eqn:pnormtinv}, establishing $y_d$ as a function
    of $y_1,\ldots,y_{d-1}$ can be replaced in form with any $y_{\ell}$, 
    indicating there are $d$ equivalent transformations, each yielding a different
    Jacobian and, for $p>1$, potentially resulting in a different calculated density.  
    Second, evaluation of geodesic distance on $\mathbb{S}_p^{d-1}$ is not 
    straightforward.  However, we have demonstrated a computationally efficient 
    upper bound on geodesic distance on $\mathbb{S}_{\infty}^{d-1}$.

### 2024-02-01 introduction.tex
changes

During the last decade or so, much work has been done in the exploration of the 
    definition and properties of an appropriate generalization of the univariate 
    GP distribution to a multivariate setting.  To mention some of the papers on 
    the topic, the work of \citep{rootzen2006} defines the generalized Pareto 
    distribution, with further analysis on these classes of distributions 
    presented in \cite{falk2008} and \cite{michel2008}.  A recent review of the 
    state of the art in multivariate peaks over threshold modelling using 
    generalized Pareto is provided in \cite{rootzen2018} while \cite{RoSeWa2018a} 
    provides insight on the theoretical properties of possible parametrizations. 
    These are use in \cite{KiRoSeWa2019} for likelihood-based models for PoT 
    estimation. A frequently used method for describing the dependence in 
    multivariate distributions is to use a copula. \cite{renard2007}, and 
    \cite{falk2019} provide successful  examples of this approach in an EVT 
    framework. \cite{ferreira2014} presents a constructive definition of the 
    Pareto process, that generalizes the GP to an infinite dimensional setting. It 
    consists of decomposing the process into independent radial and angular 
    components. Such an approach can be used in the finite dimensional case, where 
    the angular component contains the information pertaining to the dependence 
    structure of the random vector. Based on this definition, we present a novel 
    approach for modelling  the angular component with families of distributions 
    that provide flexibility and can be applied in a moderately large dimensional 
    setting.  Our focus on the angular measure is similar to that in \cite{boldi2007},
    \cite{SaNa2014} and \cite{HaCaCh2017}, that consider Bayesian non-parametric 
    approaches. Yet, our approach differs in that it is established in the 
    peaks-over-threshold regime, and avoids the problem of dealing with the so 
    called moment constraint by considering a constructive definition of the 
    multivariate GP, based on the infinity norm. The approach proposed in this paper
    adds to the growing literature on Bayesian models for multivariate extreme
    value analysis (see, for example, \cite{boldi2007},\cite{guillotte2011},
    \cite{SaNa2014},\cite{hanson2017}), providing a model that has strong computational
    advantages due its structural simplicitly, achieves flexibility using 
    a mixture model, imposes no moment constraints, and scales well to moderately 
    large dimensions.

@@

The approach considered in \cite{rootzen2018} focuses on the limiting
    conditional distribution $H$. An alternative approach consists of assuming
    that regular variation \cite[see, for example,][]{resnick2008extreme} holds 
    for the limiting distribution of $\bm{W}$, implying
    \[
        \lim\limits_{n\rightarrow\infty} n \prob{n^{-1} \bm{W}\in A} = 
        \mu(A),
    \]
    where $\mu$ is termed the exponent measure. $\mu$ has the homogeneity property
    $\mu(tA) = t^{-1}\mu(A)$. Letting $\rho = \|W\|_p, p>0$ 
    and $\bm{\theta} = \bm{W}/\rho$, define  $\Psi(B) = \mu(\{\bm{w} : \rho>1, 
    \bm{\theta} \in B\})$, which is referred to as the angular measure. After 
    some manipulations, we obtain that
    \begin{equation}
    \label{exponent-mu}
        \lim\limits_{r\rightarrow\infty} 
        \text{Pr}\left[\bm{\theta}\in A \rvert \rho>r\right] = 
            \frac{\Psi(A)}{\Psi({\mathbb S}_p^{d-1})} .
    \end{equation}
    Thus, a model for the exponent measure induces a model for the limiting distribution
    conditional on the observations being above a threshold defined with respect
    to their $p$-norm. The constraint that all marginals of $\mu$ correspond to a
    standard Pareto distribution leads to the so called moment constraints on $\Psi$. 
    Inference for the limiting distribution of the exceedances
    needs to account for the normalizing constant in Equation (\ref{exponent-mu}), as well
    the moment constraints. In addition, the moment constraints can impose unrealistic
    symmetries in the data distribution.

>>>>

During the last decade or so, much work has been done in the exploration of the 
    definition and properties of an appropriate generalization of the univariate 
    GP distribution to a multivariate setting.  To mention some of the papers on 
    the topic, the work of \citep{rootzen2006} defines the generalized Pareto 
    distribution, with further analysis on these classes of distributions 
    presented in \cite{falk2008} and \cite{michel2008}.  A recent review of the 
    state of the art in multivariate peaks over threshold modelling using 
    generalized Pareto is provided in \cite{rootzen2018} while \cite{RoSeWa2018a} 
    provides insight on the theoretical properties of possible parametrizations. 
    These are use in \cite{KiRoSeWa2019} for likelihood-based models for PoT 
    estimation. A frequently used method for describing the dependence in 
    multivariate distributions is to use a copula. \cite{renard2007}, and 
    \cite{falk2019} provide successful examples of this approach in an EVT 
    framework.

    \cite{ferreira2014} presents a constructive definition of the 
    Pareto process, that generalizes the GP to an infinite dimensional setting. It 
    consists of decomposing the process into independent radial and angular 
    components. Such an approach can be used in the finite dimensional case, where 
    the angular component contains the information pertaining to the dependence 
    structure of the random vector. Based on this definition, we present a novel 
    approach for modelling  the angular component with families of distributions 
    that provide flexibility and can be applied in a moderately large dimensional 
    setting.  Our focus on the angular measure is similar to that in \cite{boldi2007},
    \cite{SaNa2014} and \cite{HaCaCh2017}, that consider Bayesian non-parametric 
    approaches. Yet, our approach differs in that it is established in the 
    peaks-over-threshold regime, and avoids the problem of dealing with the so 
    called moment constraint by considering a constructive definition of the 
    multivariate GP, based on the infinity norm. The approach proposed in this paper
    adds to the growing literature on Bayesian models for multivariate extreme
    value analysis (see, for example, \cite{boldi2007},\cite{guillotte2011},
    \cite{SaNa2014},\cite{hanson2017}), providing a model that has strong computational
    advantages due its structural simplicitly, achieves flexibility using 
    a mixture model, imposes no moment constraints, and scales well to moderately 
    large dimensions.

@@

The approach considered in \cite{rootzen2018} focuses on the limiting
    conditional distribution $H$. An alternative approach consists of assuming
    that regular variation \cite[see, for example,][]{resnick2008extreme} holds 
    for the limiting distribution of $\bm{W}$, implying
    \[
        \lim\limits_{n\rightarrow\infty} n \prob{n^{-1} \bm{W}\in A} = 
        \mu(A),
    \]
    where $\mu$ is termed the exponent measure. $\mu$ has the homogeneity property
    $\mu(tA) = t^{-1}\mu(A)$. Letting $\rho = \|W\|_p, p>0$ 
    and $\bm{\theta} = \bm{W}/\rho$, define  $\Psi(B) = \mu(\{\bm{w} : \rho>1, 
    \bm{\theta} \in B\})$, which is referred to as the angular measure. After 
    some manipulations, we obtain that
    \begin{equation}
    \label{exponent-mu}
        \lim\limits_{r\rightarrow\infty} 
        \text{Pr}\left[\bm{\theta}\in A \rvert \rho>r\right] = 
            \frac{\Psi(A)}{\Psi({\mathbb S}_p^{d-1})} .
    \end{equation}
    Thus, a model for the exponent measure induces a model for the limiting distribution
    conditional on the observations being above a threshold defined with respect
    to their $p$-norm. The constraint that all marginals of $\mu$ correspond to a
    standard Pareto distribution leads to the so called moment constraints on $\Psi$. 
    Inference for the limiting distribution of the exceedances
    needs to account for the normalizing constant in Equation~\eqref{exponent-mu}, as well
    the moment constraints.

    % In addition, the moment constraints can impose unrealistic
    % symmetries in the data distribution.

### 2024-02-01 main_sbss.tex
sbss initial
changes

\documentclass[iicol,sn-basic]{sn-jnl}

\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{natbib}
\usepackage{amsthm}
\usepackage{hyperref}
%\usepackage{graphicx}
%\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{soul}
%\usepackage[ruled]{algorithm2e}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{wrapfig}
% \usepackage{setspace}
% \linespread{1.8}
\usepackage{anysize}
\marginsize{1in}{1in}{1in}{1in}

@@

\jyear{2023}
\theoremstyle{thmstyleone}

@@

@@

\title[~\hspace{6.75in}BNP inference for multivariate PoT]{Bayesian non-parametric inference for multivariate peaks-over-threshold models}

\author*{\fnm{Peter} \sur{Trubey}}\email{ptrubey@ucsc.edu}
\author{\fnm{Bruno} \sur{Sans\'{o}}}\email{bruno@soe.ucsc.edu}

\affil{\orgdiv{Department of Statistics}, \orgname{University of California Santa Cruz}, 
            \orgaddress{\street{1156 High Street}, \city{Santa Cruz},\postcode{95064}, \state{CA}, \country{USA}}}

\abstract{\input{./abstract}}
\keywords{Multivariate extremes, Peak over threshold models,
                Bayesian non-parametric models, Dirichlet process mixtures}

@@

\bibliography{./refs}

>>>>

\documentclass[11pt]{article}

\usepackage{amsmath,amssymb,amsfonts,amsthm}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{soul}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{setspace}
\linespread{1.8}

\usepackage{multirow}%
\usepackage{mathrsfs}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{anysize}
\usepackage[margin=1.75cm]{geometry}

@@

@@

\title{Bayesian Non-parametric Inference for Multivariate\\ Peaks-over-threshold Models}

% \author{Peter Trubey \& Bruno Sans\'{o}\\ Department of Statistics, UC Santa Cruz}
\author{}
\date{}

@@

\abstract{\input{./abstract_sbss}}
% \keywords{Multivariate extremes, Peak over threshold models,
%                 Bayesian non-parametric models, Dirichlet process mixtures}

@@

\bibliographystyle{plainnat}
\bibliography{./refs}
